Rates of true polar wander in convecting planets

### 2015-01-26 tpw_rate.tex
Work on scaling

Herein we perform a scaling analysis for a minimal system of a rotating, convecting mantle.

@@

The quantites $\Lambda_{i,j}$ and $\theta$ should be functions of our nondimensional parameters $\mathrm{Ra}$, $\mathrm{Fr}$, and $\Gamma$.  
We consider relatively slowly rotating bodies here ($\mathrm{Fr} << 1$), so that the rotation does not affect the style of convection.  
In this regime, therefore, we neglect the dependence on $\mathrm{Fr}$.

We support our scaling analyses with numerical simulations of 2D mantle convection using the finite element code \texttt{Aspect} \citep{kronbichler2012high}, based on the \texttt{deal.II} library \citep{dealII81}.  
\texttt{Aspect} provides modern numerical methods, flexible geometry, and powerful support for integrating user-defined modules and postprocessors.
We performed a number of simulations at different Rayleigh numbers, integrating Equation \ref{eq:liouville} at every time step.

@@

\begin{figure}
\centering
\label{misfit}
\includegraphics[width=0.9\textwidth]{figures/eigengap.pdf}
\caption{Oooga}
\end{figure}

@@

We can expand the temperature structure of the convecting planet with a set of orthonormal basis functions $X_{lmn}$, 
where $l$, $m$, and $n$ are wavenumbers.  
\begin{equation} 
T( x, y ,z, t )  = {\displaystyle \sum_{l m n}^\infty } T_{lmn}(t) X_{lmn} (x , y, z)
\label{T_series}
\end{equation}

@@

\begin{equation}
\lVert T(x,y,z, t) \rVert_\infty = \lVert {\displaystyle \sum_{l m n}^\infty } T_{lmn} (t) X_{lmn} (x , y, z) \rVert_\infty = 1
\end{equation}

@@

At high Rayleigh number we expect there to be a lengthscale which characterizes the smallest length of features in the convecting system.  
At lengthscales smaller than this diffusion will tend to wipe out heterogeneities, and there will be no power in those modes.  
Therefore we expect, to a good approximation, that we will be able to cut off the infinite sum in \ref{T_series} at some maximum wavenumber, set by the smallest lengthscale $d$:

@@

Any of a number of scaling laws can provide an estimate for the characteristic length scale of a convecting system which may depending on rheology, geometry, or density structure.
The simplest, based on boundary layer theory \citep{turcotte1967finite}, finds $d/R ~ \mathrm{Ra}^{-1/3}$, thus furnishing us with an estimate of the power in the degree-two part of the field as a function of Rayleigh number:

@@

Let $\delta$ be the perturbation to the convective moment of inertia, and let $\lambda_3 \ge \lambda_2 \ge \lambda_1$ be the eigenvalues of that moment.  
Then the rotation of the principal axes of the tensor is bounded by

@@

Furthermore, the quantity $(\lambda_i-\lambda_j)$ which appears in the denominator of these equations is precisely the same as the quantity which we estimated in the previous section to scale with $\sim \mathrm{Ra}^{-1}$.
Therefore, as the Rayleigh number increases, the characteristic gap between the eigenvalues of the convective moment becomes smaller.  
Additionally, the timescale of fluctuations in these values goes down. 
Overall, this makes the principle axes of high Rayleigh number systems much less stable.
This is consistent with the result of \citet{richards1997explanation}.

\begin{figure}
\centering
\label{misfit}
\includegraphics[width=0.9\textwidth]{figures/misfit.pdf}
\caption{Top: }
\end{figure}

Figure \ref{misfit} demonstrates the importance of this ``eigengap'' for the rates of true polar wander.

>>>>

Herein we perform a scaling analysis of rates of TPW for a minimal system of a rotating, convecting mantle.

@@

The quantites $\Lambda_{i,j}$ and $\theta$ should be functions of our nondimensional parameters $\mathrm{Ra}$, $\mathrm{Fr}$, and $\Gamma$.  
We consider relatively slowly rotating bodies here ($\mathrm{Fr} \ll 1$), so that the rotation does not affect the style of convection.  
In this regime, therefore, we neglect the dependence on $\mathrm{Fr}$.

We support our scaling analyses with numerical simulations of 2D mantle convection using the finite element code \texttt{Aspect} \citep{kronbichler2012high}, based on the \texttt{deal.II} library \citep{dealII81}.  
\texttt{Aspect} provides modern numerical methods, flexible geometry, and powerful support for integrating user-defined modules and postprocessors.
We performed a number of simulations at different Rayleigh numbers, integrating Equation \ref{eq:liouville} in time.

@@

\begin{figure}
\centering
\label{misfit}
\includegraphics[width=0.9\textwidth]{figures/eigengap.pdf}
\caption{ Left: Time series of the normalized difference between moments $\Lambda_{12}$ for convection in a 2D annulus at several different Rayleigh numbers.  As the Rayliehg number increases, the average value of the relative moment decreases due to less low-degree coherence in the temperature structure.  Right:  average value of $\Lambda_{12}$ for the different Rayleigh numbers.  Also shown is a line with slope $\mathrm{Ra}^{-2/3}$, which is predicted from the scaling analysis (the exponent is $-2/3$ instead of $-1$ due to the reduced dimensionality of the simulations).}
\end{figure}

@@

We can expand the temperature structure of the convecting planet with a set of orthonormal basis functions $R_n Y_{lm}$, 
where $Y_{lm}$ are spherical harmonics, $R_n$ are some set of radial polynomials.  
\begin{equation} 
T( r , \theta, \phi, t )  = {\displaystyle \sum_{n=0}^\infty \sum_{l=0}^\infty \sum_{m=-l}^{l} } T_{lmn}(t) R_n(r) Y_{lm} (\theta , \phi)
\label{T_series}
\end{equation}

@@

\begin{equation}
\lVert T(x,y,z, t) \rVert_\infty = \lVert {\displaystyle \sum_{n=0}^\infty \sum_{l=0}^\infty \sum_{m=-l}^{l} } T_{lmn}(t) R_n(r) Y_{lm} (\theta , \phi) \rVert_\infty = 1
\end{equation}

@@

At low Rayleigh number we expect the spectrum of the temperature field to be dominated by only a few low-degree modes (possibly, but not necessarily degree-two).

At high Rayleigh number we expect there to be a lengthscale which characterizes the smallest length of features in the convecting system.  
At lengthscales smaller than this diffusion will tend to wipe out heterogeneities, and there will be no power in those modes.  
Therefore we expect, to a good approximation, that we will be able to cut off the infinite sum in Equation \ref{T_series} at some maximum wavenumber, set by the smallest lengthscale $d$:

@@

Any of a number of scaling laws can provide an estimate for the characteristic length scale of a convecting system which may depending on rheology, geometry, or density structure.
The simplest, based on boundary layer theory \citep{turcotte1967finite}, finds $d/R \sim \mathrm{Ra}^{-1/3}$, thus furnishing us with an estimate of the power in the degree-two part of the field as a function of Rayleigh number:

@@

Let $\delta$ be the size of the perturbation to the convective moment of inertia tensor, and let $\lambda_3 \ge \lambda_2 \ge \lambda_1$ be the eigenvalues of that tensor.  
Then the rotation of the principal axes of the tensor is bounded by

@@

Furthermore, the quantity $(\lambda_i-\lambda_j)$ which appears in the denominator of these equations is precisely the same as the quantity which we estimated in the previous section to scale with $\sim \mathrm{Ra}^{-1}$.
Therefore, as the Rayleigh number increases, the characteristic gap between the eigenvalues of the convective moment becomes smaller.  
Additionally, the timescale of fluctuations in these values goes down. 
Overall, this makes the principle axes of high Rayleigh number systems much less stable.
This is consistent with the result of \citet{richards1999polar}.

In the limit that the eigengap becomes zero, the rotation of the pinciple axes can be arbitrary. 
This essentially corresponds to the hypothesized ``inertial interchange true polar wander'' \citep{kirschvink1997evidence}.

\begin{figure}
\centering
\label{misfit}
\includegraphics[width=0.9\textwidth]{figures/misfit.pdf}
\caption{Top: Time series of principle moments for 2D annular convection at $\mathrm{Ra}\sim10^8$.  Bottom: Time series of spin axis and mismatch angle $\theta$.  When the two moments are close to each other (small eigengap), the mismatch angle becomes large, and the rate of polar wander is significantly larger.}
\end{figure}
